The Impact of Malaria Episodes and Treatment Regimens on Adverse Pregnancy Outcomes in Ugandan Women

### 2025-03-30 code/analysis-code/Statistical_Analysis.qmd
Project

# Exploratory Data Analysis

```{r}
# Load required packages
library(ggplot2)

# Create a histogram of total malaria episodes
ggplot(data = promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Malaria Episodes",
    x = "Total Malaria Episodes",
    y = "Count"
  )

@@

```{r}
library(ggplot2)
library(dplyr)
library(here)

@@

```{r}
# Load required packages
library(here)
library(dplyr)
library(ggplot2)
library(corrplot)

@@

# Define the variables for the demographics and baseline characteristics table.
# Adjust or add variables based on your dataset.
vars <- c("age_at_enrollment_years", 
          "gestational_age_at_enrollment_weeks", 
          "education_level", 
          "gravidity", 
          "parity")

# Optionally, you can stratify by a key categorical variable such as study_arm.
# If you wish to present separate summaries for each study arm, set the stratification variable.
strata <- "study_arm"

@@

```{r}
# Load required packages
library(tibble)
library(knitr)
library(kableExtra)

@@

```{r}
library(ggplot2)
library(dplyr)
library(here)

# Create a stacked bar plot for adverse outcomes by study arm
ggplot(promo_data_clean, aes(x = study_arm, fill = factor(adverse_outcome))) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Adverse Outcomes by Treatment Arm",
    x = "Treatment Arm",
    y = "Percentage",
    fill = "Adverse Outcome\n(0 = No, 1 = Yes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

>>>>

```{r}
colnames(promo_data_clean)
```

# Exploratory Data Analysis
```{r}
# Create a histogram of total malaria episodes
ggplot(data = promo_data_clean, aes(x = total_malaria_episodes)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Total Malaria Episodes",
    x = "Total Malaria Episodes",
    y = "Count"
  )

@@

```{r}
# Read in the cleaned data
promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

@@

```{r}
# Read in the cleaned data
promo_data_clean <- read.csv(here("data", "clean", "PROMO_Data_clean.csv"))

@@

# Define the variables for the demographics and baseline characteristics table.
vars <- c("age_at_enrollment_years", 
          "gestational_age_at_enrollment_weeks", 
          "education_level", 
          "gravidity", 
          "parity")

strata <- "study_arm"

@@

```{r}

@@

```{r}
table(promo_data_clean$stillbirth_bin, useNA = "always")
table(promo_data_clean$stillbirth_bin, useNA = "always")
table(promo_data_clean$stillbirth)

@@

## Results

```{r}
# --- Load libraries ---
library(dplyr)
library(broom)
library(logistf)
library(knitr)
library(kableExtra)
library(tibble)

# --- Data preparation ---
promo_data_clean <- promo_data_clean %>%
  mutate(
    preterm_birth = ifelse(gestational_age_at_delivery_weeks < 37, 1, 0),
    stillbirth_bin = ifelse(stillbirth == "Yes", 1, 0)
  )

# --- Parsimonious Logistic Regression for Preterm Birth ---
model_preterm_pars <- glm(
  preterm_birth ~ total_malaria_episodes_during_pregnancy +
    age_at_enrollment_years + gravidity,
  data = promo_data_clean,
  family = binomial(link = "logit")
)

# --- Parsimonious Firth Logistic Regression for Stillbirth ---
model_stillbirth_pars <- logistf(
  stillbirth_bin ~ total_malaria_episodes_during_pregnancy +
    age_at_enrollment_years + gravidity,
  data = promo_data_clean
)

# --- Extract Preterm results ---
summary_preterm_pars <- tidy(model_preterm_pars, exponentiate = TRUE, conf.int = TRUE)

# --- Create Preterm Table ---
table_preterm <- summary_preterm_pars %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  mutate(term = recode(term,
                       "(Intercept)" = "Intercept",
                       "total_malaria_episodes_during_pregnancy" = "Total malaria episodes during pregnancy",
                       "age_at_enrollment_years" = "Age at enrollment (years)",
                       "gravidity" = "Gravidity"
                       ))

kable(table_preterm,
      caption = "**Table 4a. Parsimonious Logistic Regression for Preterm Birth (N = 782)**",
      col.names = c("Exposure and Adjustment Variables", "Adjusted Odds Ratio (aOR)", "95% CI (Lower)", "95% CI (Upper)", "p-value")) %>%
  kable_styling(full_width = FALSE, position = "center")

# --- Create Stillbirth Table ---
table_stillbirth <- data.frame(
  `Exposure and Adjustment Variables` = c(
    "Total malaria episodes during pregnancy",
    "Age at enrollment (years)",
    "Gravidity"
  ),
  `Adjusted Odds Ratio (aOR)` = round(exp(model_stillbirth_pars$coefficients[-1]), 2),
  `95% CI (Lower)` = round(exp(model_stillbirth_pars$ci.lower[-1]), 2),
  `95% CI (Upper)` = round(exp(model_stillbirth_pars$ci.upper[-1]), 2),
  `p-value` = signif(model_stillbirth_pars$prob[-1], 3)
)

kable(table_stillbirth,
      caption = "**Table 4b. Parsimonious Firth Logistic Regression for Stillbirth (N = 782)**") %>%
  kable_styling(full_width = FALSE, position = "center")

# --- Poisson Regression Model (Malaria Episode Rate by Treatment Arm) ---
model_poisson <- glm(
  total_malaria_episodes_during_pregnancy ~ dp_treatment + age_at_enrollment_years + gravidity + education_level,
  data = promo_data_clean,
  family = poisson(link = "log")
)

summary_poisson <- tidy(model_poisson, exponentiate = TRUE, conf.int = TRUE)

# --- Create Poisson Table ---
table_poisson <- summary_poisson %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  mutate(term = recode(term,
                       "(Intercept)" = "Intercept",
                       "dp_treatment" = "DP vs SP Treatment Arm",
                       "age_at_enrollment_years" = "Age at enrollment (years)",
                       "gravidity" = "Gravidity",
                       "education_levelSecondary" = "Education: Secondary",
                       "education_levelTertiary" = "Education: Tertiary",
                       "education_levelUniversity" = "Education: University"
                       ))

kable(table_poisson,
      caption = "**Table 4c. Poisson Regression – Malaria Episode Incidence by Treatment Arm (N = 782)**") %>%
  kable_styling(full_width = FALSE, position = "center")

# --- Create Summary Table ---
summary_results <- tribble(
  ~Outcome, ~Exposure, ~Effect_Measure, ~Estimate, ~`95% CI`, ~`p-value`,
  "Preterm Birth", "Total malaria episodes during pregnancy", "Adjusted Odds Ratio (aOR)", "1.04", "0.95–1.14", "0.401",
  "Stillbirth", "Total malaria episodes during pregnancy", "Adjusted Odds Ratio (aOR)", "0.96", "0.83–1.10", "0.559",
  "Malaria Episode Incidence", "DP vs SP Treatment Arm", "Incidence Rate Ratio (IRR)", "1.02", "0.92–1.12", "0.756"
)

kable(summary_results,
      caption = "**Table 4d. Summary of Parsimonious Regression Analyses for Primary Outcomes**") %>%
  kable_styling(full_width = FALSE, position = "center")

### 2025-03-30 products/manuscript/Manuscript.qmd
Project

>>>>

*Table 2: Association Between Malaria Episodes and Risk of Preterm Birth*
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)

# Create cleaned table
table_preterm_pars_clean <- tribble(
  ~Variable, ~`Adjusted Odds Ratio (aOR)`, ~`95% CI Lower`, ~`95% CI Upper`, ~`p-value`,
  "Total malaria episodes during pregnancy", 1.04, 0.95, 1.14, 0.401,
  "Age at enrollment (years)", 0.99, 0.96, 1.03, 0.697,
  "Gravidity", 0.91, 0.82, 1.02, 0.106,
  "Parity", 1.17, 1.05, 1.30, 0.005,
  "Hypertension", 0.89, 0.65, 1.22, 0.463,
  "Diabetes mellitus", 1.22, 0.88, 1.69, 0.237
)

# Display table nicely
kable(table_preterm_pars_clean,
      caption = "",
      col.names = c("", "Adjusted Odds Ratio (aOR)", "95% CI Lower", "95% CI Upper", "p-value")) %>%
  kable_styling(full_width = FALSE, position = "center")

*Table 3: Association Between Malaria Episodes and Risk of Stillbirth*
```{r}
# – Stillbirth Model 
library(logistf)

# Run parsimonious stillbirth model
model_stillbirth_pars <- logistf(
  stillbirth_bin ~ total_malaria_episodes_during_pregnancy +
    age_at_enrollment_years + gravidity + parity +
    hypertension + diabetes_mellitus,
  data = promo_data_clean
)

# Extract and format results
table_stillbirth_pars <- data.frame(
  `Exposure and Adjustment Variables` = names(model_stillbirth_pars$coefficients),
  `Adjusted Odds Ratio (aOR)` = round(exp(model_stillbirth_pars$coefficients), 2),
  `95% CI Lower` = round(exp(model_stillbirth_pars$ci.lower), 2),
  `95% CI Upper` = round(exp(model_stillbirth_pars$ci.upper), 2),
  `p-value` = signif(model_stillbirth_pars$prob, 3)
)

# Display in manuscript-ready table
kable(table_stillbirth_pars,
      caption = "**") %>%
  kable_styling(full_width = FALSE, position = "center")

*Association between Malaria Episodes and Adverse Birth Outcomes*
In adjusted logistic regression analysis (Table 4a), the number of malaria episodes during pregnancy was not significantly associated with preterm birth (aOR: 1.04; 95% CI: 0.95–1.14; p = 0.401). Similarly, the penalized Firth logistic regression model showed no significant association between malaria episodes and stillbirth (aOR: 0.96; 95% CI: 0.83–1.10; p = 0.559) (Table 4b). No significant associations were observed with age at enrollment, gravidity, or hypertension for either outcome.

*Effect of Treatment Arm on Malaria Episode Incidence*
A Poisson regression model assessing malaria episode incidence by treatment arm found no significant difference between the DP and SP treatment groups (IRR: 1.02; 95% CI: 0.92–1.12; p = 0.756) after adjusting for age, gravidity, and education level (Table 4c).

A summary of these primary outcome analyses is presented in Table 4.
